\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-10-06 paper.tex
Updates from Overleaf

\linenumbers
\title{\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs}

@@

\setlength{\skip\footins}{1em}
\setlength{\footnotemargin}{0.5em}
\renewcommand{\thefootnote}{\arabic{footnote}}
\footnotetext[1]{Big Data Institute, Li Ka Shing Centre for Health
Information and Discovery, University of Oxford, UK}
\footnotetext[2]{School of Mathematics and Statistics, University of Glasgow, UK}
\footnotetext[3]{Department of Statistics, University of Oxford, UK}
\footnotetext[4]{School of Mathematics, Statistics and Physics, Newcastle University, UK}
\footnotetext[5]{Department of Statistics, University of Warwick, UK}
\footnotetext[6]{The Roslin Institute and Royal (Dick) School of Veterinary Studies, University of Edinburgh, UK}
\footnotetext[7]{Broad Institute of MIT and Harvard, Cambridge, USA}
\footnotetext[8]{Department of Genetics, Stanford University School of Medicine, Stanford, USA}
\renewcommand{\thefootnote}{\fnsymbol{footnote}}
\footnotetext[1]{Shared second authorship, listed alphabetically}
\footnotetext[2]{Correspondence: jerome.kelleher@bdi.ox.ac.uk}

@@

\section*{Introduction}
Estimating the genetic genealogy of a set of genome sequences
under the influence of recombination,
usually known as an Ancestral Recombination Graph (ARG), is a long-standing
goal in genetics.
% TODO say this better.
From a high level, an ARG describes the different paths of genetic inheritance
caused by recombination, and encodes a sequence of correlated genealogical
trees along the genome.
Recent breakthroughs
in large-scale inference
methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,
schaefer2021ancestral,wohns2022unified,zhang2023biobank,zhan2023towards}
and data representation~\citep{kelleher2016efficient,kelleher2018efficient}
have raised the realistic prospect of ARG-based analysis becoming a standard part
of the population and statistical genetics toolkit~\citep{hejase2020summary}.
Applications using inferred ARGs as input have begun to
appear~\citep{osmond2021estimating,fan2022genealogical,hejase2022deep,zhang2023biobank,
nowbandegani2023extremely}
and many more are sure to
follow~\citep{harris2019database,harris2023using}.

@@

The word ``ARG'' is now widely used to refer to genetic genealogies in
general~\citep[MORE][]{mathieson2020ancestry}, including
varied outputs from modern simulation and
inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree,
kelleher2019inferring, speidel2019method, baumdicker2021efficient, zhang2023biobank}.
This broad usage means there is disagreement on the
exact object being defined. This causes problems, for example,
when evaluating inferred ARGs, and even when defining
what the criteria for success should be.
It also means there is no shared format for interchanging ARG data between
programs, resulting in the assorted negative consequences with which
geneticists are only too familiar~\citep{excoffier2006computer,gower2022demes}.
Perhaps most importantly, the direct linkage to
the original stochastic process definition leads to an
overly narrow interpretation of ARGs, which is significantly
out of step with modern methods and large-scale datasets, and
can lead to missed opportunities in their application and usage.
Such problems go much deeper than pedantic discussion over terminology;
they directly inform what we should be attempting to estimate, how we
assess success, and the expected limitations on the precision
and accuracy of our inferences.
% Thanks Gideon and Alex!
If not addressed, these problems will
impede progress on applying genetic genealogical
information to understand the biological world.

@@

% What do we do?
In this paper we attempt to address these problems by
providing a precise definition of an ARG that is general enough to encompass
the output of popular simulation and inference methods.
We propose a definition of genetic genealogy in terms of specific genomes
and their intervals of genetic inheritance, and refer to this as
a ``genome ARG'', or ``gARG''.
This contrasts with the classical Griffiths approach,
based on recording evolutionary events, which we call
an ``event ARG'', or ``eARG''.
We show that the gARG encoding has a range of advantages over the
eARG approach, being flexible, efficient and, in particular,
allowing us to systematically encode uncertainty about the
time ordering of recombination events.

@@

The common feature of these definitions is that they are expressed
in terms of evolutionary \emph{events}, and so for the purposes of
discussion we can label
this prevailing interpretation as an ``Event ARG'', or eARG.
(Many definitions also emphasise
the necessity of exhaustively capturing \emph{all} events, which we
examine in later sections.)
Another common feature of these definitions is that they do not explicitly
discuss the mechanism by which genealogical trees vary along the genome.
The topology of an eARG defines the common ancestor and
recombination events that occured,
but the graph structure alone is not sufficient to define the local trees,
or equivalently, how ``ancestral material'' (Section XXX) flows along
the edges of the ARG.
This is usually assumed to follow the mechanism described by
Griffiths and colleagues in which we have
two different ``types'' of node in the graph:
common ancestor nodes in which the inbound lineages are merged into a
single ancestral lineage with one parent, and recombination
nodes in which a single lineage is split into two independent
ancestral lineages.
Recombination nodes are then
annotated with
the corresponding crossover breakpoint~\citep{griffiths1996ancestral}.
Fig.~\ref{fig-arg-data-structure} shows an example of a classical
eARG with three samples (\noderef{a}, \noderef{b}, and \noderef{c})
and a single recombination event, along with an example concrete
encoding of the data.  There is a single recombination event
at node \noderef{d} with a breakpoint at position $x$. We
assume that \noderef{d} inherits genetic material to the
left of $x$ from \noderef{e} and to the right of $x$ from \noderef{f}.
As shown in Fig.~\ref{fig-arg-data-structure}B,
the local trees are embedded in the graph, and can be reconstructed
in essentially the same manner as outlined the previous section for
gARGs. The only difference is that decisions about which outbound
edge to follow during the rootward traversal are taken at recombination
nodes, following
either the common convention that the order of parents
is significant~\citep[e.g.][]{griffiths1991two},
or by incorporating additional information to
distinguish left and right parents
into the
encoding~\citep[e.g.][]{gusfield2014recombinatorics,ignatieva2021kwarg}.

@@

While the additional generality provided by the gARG representation
is useful, it is in reality a small change that does not
lead to significantly different time or space complexity in computational
tasks.
The vitally important and fundamental distinction between
a gARG and an eARG is that gARGs
allow us to directly describe
genetic inheritance in multiple non-contiguous genome intervals
between an ancestor and a descendant, which cannot be done
by specifying recombination events and breakpoints.
% FIXME this sentence is problematic. We should recast to talk about
% the outcomes of events, not talk about simulateousness (which will
% confuse people
Thus, we have the means of encoding the \emph{effects} of
multiple evolutionary events, and to
systematically describe the uncertainty around their
ordering (see Section XXX).
As we see in the following sections, freeing ourselves from the requirement
of recording specific events and
the additional generality of the gARG encoding
has many advantages. It also raises questions about what our goals
for inferring ARGs should be, and how we evaluate the success
of such inferences.

@@

\section*{ARGs and local trees}
% TODO mention Yamato here? I'm sure there were some interesting points 
% about local trees in that paper
Local trees, describing the genetic genealogy of a sample
at a particular location along the genome, are
embedded within in ARG, and the relationship
between the two is subtle and important.
% An ARG is sometimes referred to as being equivalent to a set of
% local trees along the genome [TODO citation], but the connection
% between the two is subtle.
% In order for the mapping to be one-to-one, so that the original
% gARG can be exactly recovered from a set of local trees and
% their genome coordinates,
% some specific properties of the ARG and the trees are required.
Note that many authors use the term ``marginal''
trees~\citep[e.g.][]{griffiths1996ancestral,minichiello2006mapping} [more]
which we avoid due the potential confusion with the
statistical concept of marginal distributions
(which could lead to interpretation as an integrated tree over many positions).

@@

The nodes in a local tree encoded
in this way are \emph{labelled} (i.e.\ carry the identity
of the corresponding ARG node), and this is fundamentally
important to how a local tree relates to its parent ARG.
If we are given the sequence of local trees for a gARG
encoded as an oriented tree (or in any other form in which
the tree nodes are explicitly labelled with the identity of the
corresponding ARG node)
along with the genome interval covered by each tree,
then we can recover precisely the same ARG. Thus, there
is a one-to-one correspondance between an ARG and
the sequence of local trees that it encodes.

@@

\cite{shipilina2023origin} discuss the same ideas,
and note that the
``full ARG ... contains more information than the series of tree
sequences along the genome''.
Their statement that an ARG
contains more information than its local trees is true only if
the local trees are unlabeled and/or do not contain unary nodes.
However, as we have shown, if the local trees
contain labelled internal (possibly unary) nodes,
then  there is precisely as
much information in them as an ARG.
Whether such unary nodes are \emph{observable}
is an entirely separate question, of course.

@@

\section*{Levels of simplification}\label{ARG_simplification}
ARG simplification is a powerful tool, and can do much more
besides removing unreachable nodes and edges in a prospective ARG
and resolving ancestral material in an eARG.
In general, we can think of
ARG simplification as the process
of removing nodes and re-writing edges (and their inheritance annotations)
to remove various types of redundancy.
The redundancy that we are interested in
revolves around the presence of nodes that appear
``unary'' in one or more local trees
(where a node has exactly one child over a particular genomic region).
These arise in several different ways.
Nodes with one child are not a standard feature of evolutionary trees;
we are usually interested in
nodes that have at least two children, because these
are in principle detectable from mutational differences
arising in the respective subtrees.
A tree branch may correspond to many generations
of individuals
% Following feedback from Wilder, this is more likely to confuse
% people unless we properly explain
% (or indeed cell divisions)
and there is usually
very little information about these intermediates, or utility
in modelling them.
% FIXME the bridge between the next few paragraphs is a bit weird, in
% that we've just been talking about how interesting unary nodes are
% but then go on about how to remove them. Just needs a bit of smoothing
% out I think.
ARGs, however, are different, and there
can be significant information about evolutionary processes
provided by nodes that are unary in one or more local trees.
% Not sure what to do with this - should mention though.
\citet{mathieson2020ancestry} show examples in which the ARGs
contain unary nodes, and state that paths are
``implicitly passing through points representing specific individuals.''
% Full quote from fig caption:
% Genetic ancestry in the form of the ARG for a single individual. Combining
% genetic ancestry from different positions leads to a graph, incorporating all
% realized genetic ancestry paths, implicitly passing through points representing
% specific individuals.

@@

The first level of simplification that we can perform is to remove
graph topology that is invisible to the samples.
An example of such topology is a
``diamond''~\citep{rasmussen2014genome}
in which the two parent nodes of a recombination immediately
join again into a common ancestor (e.g.~\noderef{j}, \noderef{l}, \noderef{m}
and \noderef{n} in Fig.~\ref{fig-simplification}A).
Unless we are specifically
interested in the recombination event or these ancestral genomes,
there is no information in this topology and the diamond can be
replaced by a single edge. More generally, any
subgraph that is singly-connected in both the leafward and
rootward direction (a ``super-diamond'') is non-identifiable and can be
replaced by one edge. This definition includes the case
of a node that has one inbound and one outbound edge, such as
nodes \noderef{f} and \noderef{h}.
Fig.~\ref{fig-simplification}B shows the result of this type of
graph topology simplification.

Simplifying away diamonds will remove many unary nodes from the
local trees, but there can still be nodes that are unary in all
of the local trees. In particular, a node can represent a recombinant
with multiple parents in the graph but only a single child (e.g.\ node \noderef{n}
in Fig.~\ref{fig-simplification}B), or can represent a common ancestor with
multiple children in the graph but in which no coalescence takes place
in the local trees
(node \noderef{r} in Fig.~\ref{fig-simplification}B).
% point out that this is why Hudson refers to CA nodes, not coalescence nodes
Such nodes are not singly connected in the graph, but are nevertheless unary in
all of the local trees.
The operation to remove them
therefore requires knowledge not just of the graph topology but also of the
ancestral material associated with the edges.
As we see in Fig.~\ref{fig-simplification}C,
removal of recombinant nodes can produce graph nodes with
more than two parents (e.g.~node \noderef{e}); and likewise, removal of
common ancestor but non-coalescent nodes can produce graph nodes with
more than two children (e.g.~node \noderef{s}). These represent a ``stacking up'' of
the \emph{effects} of multiple evolutionary events in a single node (genome), and the
ARG no longer contains the intermediate genomes representing those events.

@@

\section*{Computational efficiency and data interchange}
ARG inference methods have been successfully applied to vast datasets,
including 500,000 genotyped humans~\citep{kelleher2019inferring,zhang2023biobank}
from UK Biobank~\citep{bycroft2018genome}
and over a million SARS-Cov-2 whole genomes~\citep{zhan2023towards}.
Even larger datasets are
available~\citep[e.g.][]{halldorsson2022sequences} % TODO more
and it seems inevitable that ARGs will be inferred for them.
The efficiency with which we can store and analyse these inferred
ARGs is therefore of great importance, and a key factor
in the ultimate success of downstream ARG-based methods.
Another key factor in how important ARG-based methods become
across the wide range of predicted application
% TODO more?
areas~\citep{rasmussen2014genome,harris2019database,
hejase2020summary,schaefer2021ancestral,
fan2022genealogical,wohns2022unified,harris2023using,
nowbandegani2023extremely}
is how well inference methods and downstream analysis applications
interoperate. Without a well-defined and efficient shared format
for ARG data interchange,
users will suffer from problems of converting data from one programs
output to another, an unfortunate hallmark of population genetics
software for many years~\citep{excoffier2006computer}.
While current methods tend to tightly couple downstream analysis
of the inferred ARG with the inference
itself within the same software package,
this approach scales poorly in terms of software development
effort, and is ultimately not compatible with the widespread use
of ARGs for routine data analysis, and a healthy and diverse
software ecosystem.

@@

This view of ARGs,
decoupled from generative models and
without the hard requirement
of complete precision on all historical events, may clarify inference goals and improve
methods for evaluation.
In most cases,
ARG inference is evaluated by simulating data from a known ground truth ARG, and comparing this to the inferred version via pairwise comparison of local trees along the genome, for example
using tree distance
metrics~\citep[e.g.][]{robinson1981comparison,kendall2016mapping},
as described by \citet{kuhner2015assessing}.
% While this provides valuable insights into the method's performance,
% it is not clear how well these tree distance metrics reflect
% performance on real data.
In comparing tree-by-tree along the genome, the effects of recombination
are only incorporated in an indirect manner through the correlations
between the local trees, rather than through directly taking into account
the persistence of nodes and edges across multiple trees.
Tree distance metrics usually have $O(n^2)$ time complexity and therefore cannot be
applied to the very large sample sizes currently of interest.
A recent trend has been to move away from such tree distance-based approaches and to examine more properties of the inferred ARGs.
\citet{brandt2021evaluation} focussed on the
distributions of pairwise MRCA times; \citet{deng2021distribution} derived an approximation to the distribution of the 
waiting distances between local trees; \citet{ignatieva2023distribution}
derived the distribution of the genomic span of an edge and that of a
clade of samples. In each case, simulation studies demonstrated 
substantial differences between these quantities in simulated and reconstructed ARGs,
that were not captured using tree-by-tree comparisons.
% Not sure this is the right place for this sentence
These evaluations have been based on
ground truth data from highly idealised simulations, with sample sizes limited to at most a few thousand (typically much fewer).
The effects of the richness of real data
on modern biobank-scale or multi-generational pedigree datasets
are almost entirely unknown (beyond studies using very simplistic error models).
A suite of ARG evaluation metrics, that take into account more of the
global topology and can be applied to large ARGs, would be a
valuable addition to the field.

Interest in ARG inference methods and downstream applications is 
burgeoning, with exciting developments arriving at ever-increasing pace.
Without agreement on basic terminology and some standardisation
on data formats, however, the ARG revolution may falter.
Here, we have advocated
for decoupling the data structures used to represent ARGs
from generative processes, and have shown that the 
gARG encoding enables flexible description of 
contemporary datasets with varying levels of precision and integral
handling of some important forms of temporal uncertainty.
For ARG based methods to achieve mainstream status, we require
a rich supporting software ecosystem. 
Ideally, this would comprise a wide range of different 
inference methods specialised to different organisms, types 
and scales of data and 
% I'm thinking about things like Espalier specifically wanting
% to characterise the recombs, vs Relate mainly wanting to 
% get good local trees
specific inference goals, sharing a common ARG interchange format. 
The output of these diverse inference methods could then be 
processed by numerous downstream applications, 
enabling users to freely compare the outputs of different 
inference methods on their data, and the result of different
analyses on these inferences.
Earlier efforts to standardise ARG interchange shared 
this vision, 
but did not succeed~\citep{cardona2008extended,mcgill2013graphml}.
Today, however, we have the benefit of a pre-existing ARG
interchange format, already widely used and accompanied 
by high-quality, high-performance software (see Section XXX for details). 
Adopting the \texttt{tskit} library 
as a community standard,
utilising its rich suite of operations,
would substantially accelerate progress, and help 
establish ARG methods as mainstream genetics.

@@

The goal of this historical overview is to illustrate that the meaning of the term ``ARG" now strongly
depends on the context in which it is used, and can mean both the
stochastic process that generates genealogies in the presence of
recombination~\citep[e.g.][]{nordborg2000linkage,birkner2013ancestral,
wilton2015smc,griffiths2016coalescent},
as well as, more commonly, the concrete realisation of ancestry from a
process~\citep[e.g.][]{gusfield2014recombinatorics,mathieson2020ancestry,brandt2021evaluation}.
\cite{minichiello2006mapping} explicitly distinguished the ARG as a
stochastic process from the ARG as a data structure. We have followed them in
using the term in the latter sense, to mean a specific realised ancestral history.

>>>>

\title{\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs}

@@

\setlength{\skip\footins}{1em}
\setlength{\footnotemargin}{0.5em}
\renewcommand{\thefootnote}{\arabic{footnote}}
\footnotetext[1]{Big Data Institute, Li Ka Shing Centre for Health
Information and Discovery, University of Oxford, UK}
\footnotetext[2]{School of Mathematics and Statistics, University of Glasgow, UK}
\footnotetext[3]{Department of Statistics, University of Oxford, UK}
\footnotetext[4]{School of Mathematics, Statistics and Physics, Newcastle University, UK}
\footnotetext[5]{Department of Statistics, University of Warwick, UK}
\footnotetext[6]{The Roslin Institute and Royal (Dick) School of Veterinary Studies, University of Edinburgh, UK}
\footnotetext[7]{Broad Institute of MIT and Harvard, Cambridge, USA}
\footnotetext[8]{Department of Genetics, Stanford University School of Medicine, Stanford, USA}
\renewcommand{\thefootnote}{\fnsymbol{footnote}}
\footnotetext[1]{Joint second author, listed alphabetically}
\footnotetext[2]{Correspondence: jerome.kelleher@bdi.ox.ac.uk}

@@

\linenumbers
\section*{Introduction}
Estimating the genetic genealogy of a set of genome sequences
under the influence of recombination,
usually known as an Ancestral Recombination Graph (ARG), is a long-standing
goal in genetics.
% TODO say this better.
From a high level, an ARG describes the different paths of genetic inheritance
caused by recombination, and encodes a sequence of correlated genealogical
trees along the genome.
Recent breakthroughs
in large-scale inference
methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,
schaefer2021ancestral,wohns2022unified,zhang2023biobank,zhan2023towards}
and data representation~\citep{kelleher2016efficient,kelleher2018efficient}
have raised the realistic prospect of ARG-based analysis becoming a standard part
of the population and statistical genetics toolkit~\citep{hejase2020summary}.
Applications using inferred ARGs as input have begun to
appear~\citep{osmond2021estimating,fan2022genealogical,hejase2022deep,zhang2023biobank,
nowbandegani2023extremely,ignatieva2023distribution}
and many more are sure to
follow~\citep{harris2019database,harris2023using}.

@@

The word ``ARG'' is now widely used to refer to genetic genealogies in
general~\citep[][AND OTHERS]{mathieson2020ancestry}, including
varied outputs from modern simulation and
inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree,
kelleher2019inferring, speidel2019method, baumdicker2021efficient, zhang2023biobank}.
This broad usage means there is disagreement on the
exact object being defined. This causes problems, for example,
when evaluating inferred ARGs, and even when defining
what the criteria for success should be.
It also means there is no shared format for interchanging ARG data between
programs, resulting in the assorted negative consequences with which
geneticists are only too familiar~\citep{excoffier2006computer,gower2022demes}.
Perhaps most importantly, the direct linkage to
the original stochastic process definition leads to an
overly narrow interpretation of ARGs, which is significantly
out of step with modern methods and large-scale datasets, and
can lead to missed opportunities in their application and usage.
Such problems go much deeper than pedantic discussion over terminology;
they directly inform what we should be attempting to estimate, how we
assess success, and the expected limitations on the precision
and accuracy of our inferences.
% Thanks Gideon and Alex!
If not addressed, these problems will
impede progress on applying genetic genealogical
information to understand the biological world.

@@

% What do we do?
In this paper, we attempt to address these problems by
providing a precise definition of an ARG that is general enough to encompass
the output of popular simulation and inference methods.
We propose a definition of genetic genealogy in terms of specific genomes
and their intervals of genetic inheritance, and refer to this as
a ``genome ARG'', or ``gARG''.
This contrasts with the classical Griffiths approach,
based on recording evolutionary events, which we call
an ``event ARG'', or ``eARG''.
We show that the gARG encoding has a range of advantages over the
eARG approach, being flexible, efficient and, in particular,
allowing us to systematically encode uncertainty about the
time ordering of recombination events.

@@

The common feature of these definitions is that they are expressed
in terms of evolutionary \emph{events}, and so for the purposes of
discussion we can label
this prevailing interpretation as an ``event ARG'', or eARG.
(Many definitions also emphasise
the necessity of exhaustively capturing \emph{all} events, which we
examine in later sections.)
Another common feature of these definitions is that they do not explicitly
discuss the mechanism by which genealogical trees vary along the genome.
The topology of an eARG defines the common ancestor and
recombination events that occured,
but the graph structure alone is not sufficient to define the local trees,
or equivalently, how ``ancestral material'' (Section XXX) flows along
the edges of the ARG.
This is usually assumed to follow the mechanism described by
Griffiths and colleagues in which we have
two different ``types'' of node in the graph:
common ancestor nodes in which the inbound lineages are merged into a
single ancestral lineage with one parent, and recombination
nodes in which a single lineage is split into two independent
ancestral lineages.
Recombination nodes are then
annotated with
the corresponding crossover breakpoint~\citep{griffiths1996ancestral}.
Fig.~\ref{fig-arg-data-structure} shows an example of a classical
eARG with three samples (\noderef{a}, \noderef{b}, and \noderef{c})
and a single recombination event, along with an example concrete
encoding of the data.  There is a single recombination event
at node \noderef{d} with a breakpoint at position $x$. We
assume that \noderef{d} inherits genetic material to the
left of $x$ from \noderef{e} and to the right of $x$ from \noderef{f}.
As shown in Fig.~\ref{fig-arg-data-structure}B,
the local trees are embedded in the graph, and can be reconstructed
in essentially the same manner as outlined the previous section for
gARGs. The only difference is that decisions about which outbound
edge to follow during the rootward traversal are taken at recombination
nodes, following
either the common convention that the order of parents
is significant~\citep[e.g.][]{griffiths1991two},
or by incorporating additional information to
distinguish left and right parents
into the
encoding~\citep[e.g.][]{gusfield2014recombinatorics,ignatieva2021kwarg}.

@@

While the additional generality provided by the gARG representation
is useful, it is in reality a small change that does not
lead to significantly different time or space complexity in computational
tasks.
The vitally important and fundamental distinction between
a gARG and an eARG is that gARGs
allow us to directly describe
genetic inheritance in multiple non-contiguous genome intervals
between an ancestor and a descendant, which cannot be done
by specifying recombination events and breakpoints.
% FIXME this sentence is problematic. We should recast to talk about
% the outcomes of events, not talk about simulateousness (which will
% confuse people
Thus, we have the means to encode the \emph{effects} of
multiple evolutionary events, and to
systematically describe the uncertainty around their
ordering (see Section XXX).
As we see in the following sections, freeing ourselves from the requirement
of recording specific events and
the additional generality of the gARG encoding
has many advantages. It also raises questions about what our goals
for inferring ARGs should be, and how we evaluate the success
of such inferences.

@@

\section*{ARGs and local trees}
% TODO mention Yamato here? I'm sure there were some interesting points 
% about local trees in that paper
Local trees, describing the genetic genealogy of a sample
at a particular location along the genome, are
embedded within in ARG, and the relationship
between the two is subtle and important.
% An ARG is sometimes referred to as being equivalent to a set of
% local trees along the genome [TODO citation], but the connection
% between the two is subtle.
% In order for the mapping to be one-to-one, so that the original
% gARG can be exactly recovered from a set of local trees and
% their genome coordinates,
% some specific properties of the ARG and the trees are required.
Note that many authors use the term ``marginal''
trees~\citep[e.g.][AND OTHERS]{griffiths1996ancestral,minichiello2006mapping}
which we avoid due the potential confusion with the
statistical concept of marginal distributions
(which could lead to interpretation as an integrated tree over many positions).

@@

The nodes in a local tree encoded
in this way are \emph{labelled} (i.e.\ carry the identity
of the corresponding ARG node), and this is fundamentally
important to how a local tree relates to its parent ARG.
If we are given the sequence of local trees for a gARG
encoded as an oriented tree (or in any other form in which
the tree nodes are explicitly labelled with the identity of the
corresponding ARG node)
along with the genome interval covered by each tree,
then we can recover precisely the same ARG. Thus, there
is a one-to-one correspondence between an ARG and
the sequence of local trees that it encodes.

@@

\cite{shipilina2023origin} discuss the same ideas,
and note that the
``full ARG...~contains more information than the series of tree
sequences along the genome''.
Their statement that an ARG
contains more information than its local trees is true only if
the local trees are unlabeled and/or do not contain unary nodes.
However, as we have shown, if the local trees
contain labelled internal (possibly unary) nodes,
then  there is precisely as
much information in them as an ARG.
Whether such unary nodes are \emph{observable}
is an entirely separate question, of course.

@@

\section*{Levels of simplification}\label{ARG_simplification}
ARG simplification is a powerful tool, and can do much more
besides removing unreachable nodes and edges in a prospective ARG
and resolving ancestral material in an eARG.
In general, we can think of
ARG simplification as the process
of removing nodes and re-writing edges (and their inheritance annotations)
to remove various types of redundancy.
The redundancy that we are interested in
revolves around the presence of nodes that appear
``unary'' in one or more local trees
(where a node has exactly one child over a particular genomic region).
These arise in several different ways.
Nodes with one child are not a standard feature of evolutionary trees;
we are usually interested in
nodes that have at least two children, because these
are, in principle, detectable from mutational differences
arising in the respective subtrees.
A tree branch may correspond to many generations
of individuals
% Following feedback from Wilder, this is more likely to confuse
% people unless we properly explain
% (or indeed cell divisions)
and there is usually
very little information about these intermediates, or utility
in modelling them.
% FIXME the bridge between the next few paragraphs is a bit weird, in
% that we've just been talking about how interesting unary nodes are
% but then go on about how to remove them. Just needs a bit of smoothing
% out I think.
ARGs, however, are different, and there
can be significant information about evolutionary processes
provided by nodes that are unary in one or more local trees.
% Not sure what to do with this - should mention though.
\citet{mathieson2020ancestry} show examples in which the ARGs
contain unary nodes, and state that paths are
``implicitly passing through points representing specific individuals.''
% Full quote from fig caption:
% Genetic ancestry in the form of the ARG for a single individual. Combining
% genetic ancestry from different positions leads to a graph, incorporating all
% realized genetic ancestry paths, implicitly passing through points representing
% specific individuals.

@@

The first level of simplification that we can perform is to remove
parts of the graph topology that are invisible to the samples.
An example of such topology is a
``diamond''~\citep{rasmussen2014genome}
in which the two parent nodes of a recombination immediately
join again into a common ancestor (e.g.~\noderef{j}, \noderef{l}, \noderef{m}
and \noderef{n} in Fig.~\ref{fig-simplification}A).
Unless we are specifically
interested in the recombination event or these ancestral genomes,
there is no information in this topology and the diamond can be
replaced by a single edge. More generally, any
subgraph that is singly-connected in both the leafward and
rootward direction (a ``super-diamond'') is non-identifiable and can be
replaced by one edge. This definition includes the case
of a node that has one inbound and one outbound edge, such as
nodes \noderef{f} and \noderef{h}.
Fig.~\ref{fig-simplification}B shows the result of this type of
graph topology simplification.

Simplifying away diamonds will remove many unary nodes from the
local trees, but there can still be nodes that are unary in all
of the local trees. In particular, a node can represent a recombinant
with multiple parents in the graph but only a single child (e.g.\ node \noderef{n}
in Fig.~\ref{fig-simplification}B), or can represent a common ancestor with
multiple children in the graph but in which no coalescence takes place
in the local trees
(node \noderef{r} in Fig.~\ref{fig-simplification}B).
% point out that this is why Hudson refers to CA nodes, not coalescence nodes
Such nodes are not singly connected in the graph, but are nevertheless unary in
all of the local trees.
The operation to remove them
therefore requires knowledge not just of the graph topology but also of the
ancestral material associated with the edges.
As we see in Fig.~\ref{fig-simplification}C,
removal of recombinant nodes can produce graph nodes with
more than two parents (e.g.~node \noderef{e}); and likewise, removal of
common ancestor but non-coalescent nodes can produce graph nodes with
more than two children (e.g.~node \noderef{s}). These represent merging
the \emph{effects} of multiple evolutionary events in a single node (genome), and the
ARG no longer contains the intermediate genomes representing those events.

@@

\section*{Computational efficiency and data interchange}
ARG inference methods have been successfully applied to vast datasets,
including 500,000 genotyped humans~\citep{kelleher2019inferring,zhang2023biobank}
from UK Biobank~\citep{bycroft2018genome}
and over a million SARS-Cov-2 whole genomes~\citep{zhan2023towards}.
Even larger datasets are
available~\citep[e.g.][]{halldorsson2022sequences} % TODO more
and it seems inevitable that ARGs will be inferred for them.
The efficiency with which we can store and analyse these inferred
ARGs is therefore of great importance, and a key factor
in the ultimate success of downstream ARG-based methods.
Another key factor in how important ARG-based methods become
across the wide range of possible applications
% TODO more?
areas~\citep{rasmussen2014genome,harris2019database,
hejase2020summary,schaefer2021ancestral,
fan2022genealogical,wohns2022unified,harris2023using,
nowbandegani2023extremely,ignatieva2023duration}
is how well inference methods and downstream analysis applications
interoperate. Without a well-defined and efficient shared format
for ARG data interchange,
users will suffer from problems of converting data from one programs
output to another, an unfortunate hallmark of population genetics
software for many years~\citep{excoffier2006computer}.
While current methods tend to tightly couple downstream analysis
of the inferred ARG with the inference
itself within the same software package,
this approach scales poorly in terms of software development
effort, and is ultimately not compatible with the widespread use
of ARGs for routine data analysis, and a healthy and diverse
software ecosystem.

@@

This view of ARGs,
decoupled from generative models and
without the hard requirement
of complete precision on all historical events, may clarify inference goals and improve
methods for evaluation.
In most cases,
ARG inference is evaluated by simulating data from a known ground truth ARG, and comparing this to the inferred version via pairwise comparison of local trees along the genome, for example
using tree distance
metrics~\citep[e.g.][]{robinson1981comparison,kendall2016mapping},
as described by \citet{kuhner2015assessing}.
% While this provides valuable insights into the method's performance,
% it is not clear how well these tree distance metrics reflect
% performance on real data.
In comparing tree-by-tree along the genome, the effects of recombination
are only incorporated in an indirect manner through the correlations
between the local trees, rather than through directly taking into account
the persistence of nodes and edges across multiple trees.
Tree distance metrics usually have $O(n^2)$ time complexity and therefore cannot be
applied to the very large sample sizes currently of interest.
A recent trend has been to move away from such tree distance-based approaches and to examine more properties of the inferred ARGs.
\citet{brandt2021evaluation} focussed on the
distributions of pairwise MRCA times; \citet{deng2021distribution} derived an approximation to the distribution of the 
waiting distances between local trees; \citet{ignatieva2023distribution}
derived the distribution of the genomic span of an edge and that of a
clade of samples. In each case, simulation studies demonstrated 
substantial differences between these quantities in simulated and reconstructed ARGs,
that were not captured using tree-by-tree comparisons.
% Not sure this is the right place for this sentence
These evaluations have been based on
ground truth data from highly idealised simulations, with sample sizes limited to at most a few thousand (typically much fewer).
The effects of the richness of real data
on modern biobank-scale or multi-generational pedigree datasets
are almost entirely unknown (beyond studies using very simplistic error models).
The development of ARG evaluation metrics that take into account more of the
global topology and can be applied to large ARGs would be a
valuable and timely addition to the field.

Interest in ARG inference methods and downstream applications is 
burgeoning, with exciting developments arriving at ever-increasing pace.
Without agreement on basic terminology and some standardisation
on data formats, however, the ARG revolution may falter.
Here, we have advocated
for decoupling the data structures used to represent ARGs
from generative processes, and have shown that the 
gARG encoding enables flexible description of 
contemporary datasets with varying levels of precision and integral
handling of some important forms of temporal uncertainty.
For ARG-based methods to achieve mainstream status, we require
a rich supporting software ecosystem. 
Ideally, this would comprise a wide range of different 
inference methods specialised to different organisms, types 
and scales of data and 
% I'm thinking about things like Espalier specifically wanting
% to characterise the recombs, vs Relate mainly wanting to 
% get good local trees
specific inference goals, sharing a common ARG interchange format. 
The output of these diverse inference methods could then be 
processed by numerous downstream applications, 
enabling users to freely compare the outputs of different 
inference methods on their data, and the result of different
analyses on these inferences.
Earlier efforts to standardise ARG interchange shared 
this vision, 
but did not succeed~\citep{cardona2008extended,mcgill2013graphml}.
Today, however, we have the benefit of a pre-existing ARG
interchange format, already widely used and accompanied 
by high-quality, high-performance software (see Section XXX for details). 
Adopting the \texttt{tskit} library 
as a community standard,
utilising its rich suite of operations,
would substantially accelerate progress, and help 
establish ARG methods as mainstream genetics.

@@

The goal of this historical overview is to illustrate that the meaning of the term ``ARG" now strongly
depends on the context in which it is used, and can mean both the
stochastic process that generates genealogies in the presence of
recombination~\citep[e.g.][]{nordborg2000linkage,birkner2013ancestral,
wilton2015smc,griffiths2016coalescent},
as well as, more commonly, the concrete realisation of ancestry from a
process~\citep[e.g.][]{gusfield2014recombinatorics,mathieson2020ancestry,brandt2021evaluation}.
\cite{minichiello2006mapping} explicitly distinguished the ARG as a
stochastic process from the ARG as a data structure. We have followed them and
used the term in the latter sense, to mean a specific realised ancestral history.
